---
title: "Derive information with dplyr"
author:
  - [TEAM NAME GOES HERE]
output: 
  html_document:
    df_print: paged
    highlight: textmate 
    theme: readable
    code_folding: hide
    toc: true
    toc_float: true
    fig_width: 9
    fig_height: 6
---

# Overview

NHANES (The National Health and Nutrition Examination Survey) was designed to assess the health and nutritional status of the US population and is conducted by the National Center for Health Statistics of the Centers for Disease Control and Prevention. Since 1999-2000, NHANES has been conducted in two-year cycles. For each cycle, potential participants are identified through stratified, multistage probability sampling of the non-institutionalized US population. In this set of exercises, we will use the ten cycles conducted from 1999-2000 through 2017-2018. 

## Import

```{r, message=FALSE, warning=FALSE}

# set gloabal rmarkdown options
# echo = TRUE     --> show code blocks in markdown doc
# message = FALSE --> suppress package startup and other messages
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

# Load packages 
library(tidyverse)  # includes dplyr - filter / select
library(gt)         # for tables

nhanes <- read_rds('data/NHANES_analysis_post_exclusions.rds')

# trying to print all rows will make df_print: paged cause a pandoc error.
# don't modify the line below! It may break knitr.
nhanes[1:10, ]
  
```


## Data dictionary

Review this briefly and use it as a reference to engage with the exercises below.

```{r}

# A data dictionary is more comprehensive than these two columns,
# but these two columns are the core components of a data dictionary.
# I highly recommend you only engage with data when there is a 
# data dictionary available. If a collaborator cannot share one of
# these with you when they ask you to analyze your data, then they 
# should not be your collaborator.

nhanes_descr <- c(
  "seqn"              = "SP identifier",
  "exam"              = "NHANES exam year",
  "psu"               = "primary sampling unit",
  "strata"            = "survey strata",
  "wts_mec_2yr"       = "survey weights",
  "age"               = "SP age, years",
  "age_group"         = "SP age group, years",
  "sex"               = "SP sex",
  "race_ethnicity"    = "SP race and/or ethnicity",
  "education"         = "SP education",
  "income_hh"         = "SP household income",
  "bp_sys_mmhg"       = "SP systolic blood pressure, mm Hg",
  "bp_dia_mmhg"       = "SP diastolic blood pressure, mm Hg",
  "bp_controlled"     = "Did SP have controlled BP? (<140/90 mm Hg)",
  "acr_mgg"           = "SP albumin-to-creatinine ratio, mg/g",
  "albuminuria"       = "Did SP have albuminuria? (ACR > 30 mg/g)",
  "chol_hdl_mgdl"     = "SP HDL-cholesterol, mg/dl",
  "chol_total_mgdl"   = "SP total cholesterol, mg/dl",
  "health_insurance"  = "SP health insurance status",
  "bp_high_aware"     = "SP ever told by Dr: 'you have high blood pressure'?",
  "bp_meds"           = "SP currently using antihypertensive medication?",
  "hc_usual_facility" = "SP has a usual healthcare facility?",
  "hc_visit_1yr"      = "SP visited their healthcare facility last year?"
)

# abbreviations need to be placed in the footnote of the table
abbrevs <- c("SP = survey participant", 
  "BP = blood pressure", 
  "HDL = high density lipoprotein")

# the enframe function transforms a vector into a tibble,
enframe(nhanes_descr) %>% 
  gt(rowname_col = "name") %>%
  tab_stubhead(label = 'Variable name') %>% 
  cols_label(value = 'Variable description') %>% 
  cols_align('right') %>% 
  tab_source_note(source_note = 'SP = survey participant; NHANES = National Health and Nutrition Examination Survey') %>% 
  tab_header(title = 'Description of NHANES data')

```

# Pre-requisites

Before starting these exercises, you should have a good understanding of 

1. The [Derive Information with dplyr](https://rstudio.cloud/learn/primers/2.3) Primer.

1. Chapter 5 of [R for Data Science](https://r4ds.had.co.nz/transform.html)

# Exercise 1

Group the `nhanes` data by exam and sex, then apply the `summary()` function to compute the mean systolic blood pressure and standard error for the mean systolic blood pressure at each exam year, for men and women, separately. 

Notes: 

- the standard error of a vector $x$ with $n$ observations is $$\frac{\text{standard deviation of x}}{\sqrt{n}}$$

- the mean systolic blood pressure should be named `sbp_mn`

- the mean systolic blood pressure should be named `sbp_se`

```{r}

# your code goes here

# Hints: 
# 
# there is a function in dplyr called n()
# that counts the number of rows in each group.
# for example, try running:
# 
# nhanes %>% 
#   group_by(exam) %>% 
#   summarise(n = n())
# 
# how could you use this to compute standard error?

```

Your solution should look like this:

```{r}

read_rds('solutions/01_solution.rds')

```

# Exercise 2

Using the summary data you created in problem 1, create a figure to explore trends in mean systolic blood pressure. 

- Talk with your teammates about the trends you see in the figure. What do you think drove the trends you see?

- Be careful! The results we are looking at are not adjusted using the NHANES weights. Therefore, results are not nationally representative.

Notes: 

- `geom_line` expects a grouping aeshetic, specified with `aes(group = <grouping variable>)`.

- If you'd like a challenge, add standard error bars. (optional)

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}

read_rds('solutions/02_solution.rds')

```

# Exercise 3

Using the `nhanes` data, compute race-by-sex specific estimates of the prevalence of hypertension in each exam year.

Note: 

- hypertension is defined as having systolic blood pressure > 140 mm Hg, diastolic blood pressure > 90 mm Hg, or currently using antihypertensive medications.

- You will want to group by three variables.

```{r}

# Your code goes here

```


Your solution should look like this:

```{r}

read_rds('solutions/03_solution.rds')

```

# Exercise 4

Apply functions from the `gt` package to reproduce the table below.

Notes: 

- `gt` handles grouped datasets by creating a table subsection for each group.

- `fmt_percent` and `fmt_missing` are used to control formatting

- `summary_rows` is used to create the Overall summary for each sex group.

```{r}

# this will be useful
exam_years <- c(
  "1999",
  "2001",
  "2003",
  "2005",
  "2007",
  "2009",
  "2011",
  "2013",
  "2015",
  "2017"
)

# Your code goes here

```


Your solution should look like this:

```{r}

read_rds('solutions/04_solution.rds')

```


# Exercise 5

Using the dataset you created in exercise 3, compare the prevalence of hypertension between 2013 and 2017 for race-sex groups. Compute the increase (relative and absolute) in prevalence for each group.

Notes: 

I recommend using `group_by`, then `filter`, then `summarize`, then `mutate`. Last, arrange the data by `sex` and (in descending order) the absolute increase in the prevalence of hypertension.

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}

read_rds('solutions/05_solution.rds')

```


# Exercise 6

Tabulate the data you created in exercise 5, using some of the stylistic features that `gt` offers. 

Notes:

- the function used to create table headers spanning columns is `tab_spanner()`

- the function used to adjust printed column names if `cols_label()`

- the function used to color cells is `data_color()`

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}

read_rds('solutions/06_solution.rds')

```



